# Rhode Island DOH Discharge Data Analysis
title: "Todd DiLullo - Midterm Project"
author: "Todd DiLullo"
date: "2024-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(haven)
df = read_csv('midterm.csv')
n = nrow(df)

# subset 1000 rows of the data
df1 = df[sample(1:n, 1000), ]
```

```{r}
df <- df %>%
  filter(yod == 18)
```


```{r}
library(dplyr)
# Selecting Variables to work with
df <- df %>%
  select("yod", "payfix","pay_ub92","age",  
                      "sex","raceethn","provider","moa", 
                      "yoa","mod","admtype", "asource" , 
                      "preopday" ,"los", "service" , "icu","ccu",    
                      "dispub92", "payer"  ,"drg","trandb", 
                      "randbg","randbs","orr", "anes","seq",   
                      "lab","dtest", "ther","blood","phar", 
                      "other","patcon","bwght","total","tot" ,  
                      "ecodub92","b_wt","pt_state","diag_adm","ancilar" ,
                      "campus","er_fee","er_chrg","er_mode","obs_chrg",
                      "obs_hour","psycchrg","nicu_day")
```

```{r}
# Variables with missing values are 'payfix', 'preopday','raceethn', 'admtype', 'asource', 'bwght', 'ecodub92', 'diag_adm','er_mode', 'obs_hour', 'nicu_day'

# Found missing values with
missing_values <- df %>%
    summarise_all(~ sum(is.na(.)))

print(missing_values)
```

```{r}
# Remove all variables with missing values
df <- df %>%
  select_if(~ !any(is.na(.)))
```

```{r}
# The variable that records the month of admission is "MOA"
# The variable that records the month of discharge is "MOD"
```

```{r}
#Month with the most admitted patients
most_admissions_month <- df %>%
  count(moa) %>%
  arrange(desc(n)) %>%
  slice(1)

print(most_admissions_month)

```
```{r}
#Month with the most admitted male patients
most_admissions_male_month <- df %>%
 filter(sex == 1) %>%
  count(moa) %>%
  arrange(desc(n)) %>%
  slice(1)

print(most_admissions_male_month)
```
```{r}
# Month with the most teenage female patients. Check and see if need moa here too
most_female_teenage_month <- df %>%
 filter(sex == 2, age >12, age <=19) %>%
  count(moa) %>%
  arrange(desc(n)) %>%
  slice(1)

print(most_female_teenage_month)
```
```{r}
# Provider 7205 had the most female patients in October 

# Provider with the most female patients in October

female_patients_october <- df %>%
  filter(moa == 10 & sex == 2)
  
provider_counts <- female_patients_october %>%
  group_by(provider) %>%
  summarise(FemaleCount = n()) %>%
  arrange(desc(FemaleCount))


top_provider <- provider_counts[1, ]
print(top_provider)
```
```{r}
#Calculating Average Age of Females and Males
avg_age_male <- mean(df$age[df$sex == 1], na.rm = TRUE)
avg_age_female <- mean(df$age[df$sex == 2], na.rm = TRUE)

cat("Average age of males:", avg_age_male, "\n")
cat("Average age of females:", avg_age_female, "\n")

# Per results below we can see that the average age is higher for males
```
```{r}
#Finding the average age of patients by month, and the month with the highest average age. 
library(dplyr)
average_age_by_month <- df %>%
  group_by(moa) %>%
  summarize(avg_age = mean(age, na.rm = TRUE))

# Find the month with the oldest patients
oldest_month <- average_age_by_month %>%
  filter(avg_age == max(avg_age))

# Print all average ages by month
print(average_age_by_month)

# Print the oldest month
cat("The month with the oldest patients on average is:", oldest_month$moa, 
    "with an average age of", oldest_month$avg_age, "\n")
```
```{r}
# Name of the provider with the highest total charge
highest_charge_provider <- df %>%
  filter(total == max(total, na.rm = TRUE)) %>%
  select(provider, total)

# Print the provider number and the charge
cat("The provider with the highest single patient charge is Provider", highest_charge_provider$provider, 
    "with a total charge of", highest_charge_provider$total, "\n")
```

```{r}
library(dplyr)
# Name of the provider with the least total charge for teenage males

df$total <-as.numeric(df$total)

least_charge_provider <- df %>%
  filter(age >= 13 & age <= 19, sex == 1) %>%  # Teenage males (age 13-19, sex = 1)
  group_by(provider) %>%  # Group by provider
  summarize(avg_total_charge = mean(total, na.rm = TRUE)) %>%  # Calculate average total charge
  filter(avg_total_charge == min(avg_total_charge))  # Find the provider with the least charge

# Print the result ensuring provider is treated as numeric
cat("The provider with the least total charge for teenage males on average is Provider", 
    least_charge_provider$provider[1], "with an average charge of", least_charge_provider$avg_total_charge[1], "\n")
```
```{r}
#Create a Season Variable. Calculate length of stays by season. Which season has the longest length of stay on average. 

average_los_by_season <- df %>%
  mutate(season = case_when(
    moa %in% 3:5 ~ "Spring",    # March, April, May
    moa %in% 6:8 ~ "Summer",    # June, July, August
    moa %in% 9:11 ~ "Fall",     # September, October, November
    moa %in% c(12, 1, 2) ~ "Winter" # December, January, February
  )) %>%
  group_by(season) %>%
  summarize(avg_los = mean(los, na.rm = TRUE)) %>%
  arrange(desc(avg_los))  # Sort by average LOS in descending order

# Print the results
print(average_los_by_season)

# Identify the season with the longest LOS on average
cat("The season with the longest length of stay on average is:", average_los_by_season$season[1], 
    "with an average LOS of", average_los_by_season$avg_los[1], "\n")
```

```{r}
# Create Season Variable
df <- df %>%
  mutate(season = case_when(
    moa %in% 3:5 ~ "Spring",    # March, April, May
    moa %in% 6:8 ~ "Summer",    # June, July, August
    moa %in% 9:11 ~ "Fall",     # September, October, November
    moa %in% c(12, 1, 2) ~ "Winter" # December, January, February
  ))
```




```{r}

# How Much does a 20 year old male get charged for staying 1 day in Fall on average

avg_charge_fall_20_male <- df %>%
  filter(age == 20, sex == 1, season == "Fall", los == 1) %>%  # Filter based on the criteria
  summarize(avg_charge = mean(total, na.rm = TRUE))  # Calculate average total charge

# Print the result
cat("The average charge for a 20-year-old male staying 1 day in the Fall season is:", 
    round(avg_charge_fall_20_male$avg_charge, 2), "\n")
```
# Through analyzing the Rhode Island Department of Health Hospital Discharge Data, There were numerous findings related to patients ages, sex, total charges, provider charges, length of stays, and even how much a 20 year old gets charged for a day in the fall. To start we determined October of 2018 had the most number admitted patients, including the most male patients. Continuing our analyis, we found that the average patient age is higher for males with an average age of 51 years old. In all, we found that the average patient age was around 50, with juanuary having the highest average of 52 years old. One of the more prominent findings was Women and Infants had the highest total charge, with $3,402,056 in total charges. We also found that Roger Williams  had the lowest total charge for a teenage male with total charges of $10,360.44. This can also tell us that total charges in all can be relatively expensive, if a lower end is $10,000 + and the high end is more than $3,000,000. Adding to that, we also found that a 20 year old male got charged $16,511 on average, for one day of stay in the fall. Lastly we noticed that the average length of stay is around 5 days, with summer nearing the high end of the averages. Overall, we can conclude that hospital patients tend to be 50 years old, length of stays tend to be 5 days, and total charges for your stay can be very expensive. 

# Visualization

```{r}
library(tidyverse)
library(lubridate)
library(knitr)
df %>% filter(los <200, sex ==1|sex == 2) %>% ggplot()+
  geom_point(mapping = aes(x = los, y = total, color = factor(sex)))+
  labs(title = "Charges vs. Length of Stay", x = "Length of Stay (days)", y = "Total Charges", caption = "This plot shows the relationship between length of stay and total charges for patients") +
  theme_minimal()


# This plot shows us that the longer patients stay, the higher their total charges will be. There is a relatively even spread of male and female charges
```



```{r}

df <- df %>%
  filter(sex ==1|sex == 2) %>% 
  mutate(age_group = case_when(
    age < 20 ~ "Under 20",
    age >= 20 & age < 40 ~ "20-39",
    age >= 40 & age < 60 ~ "40-59",
    age >= 60 ~ "60+"
  ))

df %>%  ggplot() +
  geom_boxplot(aes(x = age_group, y = total, color = factor(sex))) +
  labs(title = "Total Charges by Age Group", x = "Age Group", y = "Total Charges", color = "Sex", caption = "This box plot shows the relationship between different age groups and total charges") +
  theme_minimal()

# This boxplot shows that as the age group decreases the totalcharges tend to increase. 
```


```{r}
total_by_provider <- df %>%
  group_by(provider) %>%
  summarize(total_charge = sum(total, na.rm = TRUE)) %>%
  arrange(desc(total_charge))

# Horizontal bar plot
ggplot(total_by_provider, aes(x = reorder(provider, total_charge), y = total_charge)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Total Charges by Provider", x = "Provider", y = "Total Charges",caption = "This plot shows the total charges by each provider of care") +
  theme_minimal()

# Here we can see that Rhode Island Hospital by far has the highest total charges then any of the other providers. 
```

```{r}
ggplot(df, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Age Distribution of Patients", x = "Age", y = "Count", caption = "This plot shows the distribution between the different age groups of patients") +
  theme_minimal()

# Here we can see that the large majority of patients are of age 0 which can be attributed to newborns, the other majority of patients are between 55-80 years old. 
```

```{r}
gender_distribution <- df %>%
  group_by(sex) %>%
  summarize(count = n())

# Bar plot
ggplot(gender_distribution, aes(x = factor(sex), y = count, fill = factor(sex))) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = c("Male", "Female")) +
  labs(title = "Gender Distribution of Patients", x = "Gender", y = "Count", caption = "This plot shows the distribution between male and female patients") +
  theme_minimal()
 
# This plot shows the distribution between the number of male and female patients, here we can see that there are more female patients then male. 
```

```{r}
df %>%
ggplot (aes(x = season, y = total, fill = season)) +
geom_boxplot() +
labs( title = "Distribution of Charges by Season", x = "Season", y = "Total Charges", caption = "This Box plot shows the spread of charges across seasons.") +
  theme_minimal()

# Here we can see that the fall season has the highest total charges, with a few outliars in summer and winter. 
```

```{r}
avg_los_by_provider <- df %>%
  group_by(provider) %>%
  summarize(avg_los = mean(los, na.rm = TRUE)) %>%
  arrange(desc(avg_los))

# Plot
ggplot(avg_los_by_provider, aes(x = reorder(provider, avg_los), y= avg_los)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  coord_flip() +
  labs(title = "Average Length of Stay by Provider",x = "Provider",y = "Average Length of Stay (days)",caption = "This plot Shows the variation in average LOS across providers.") +
  theme_minimal()

# Here we can see that Bradley Hospital has the longest average length of stay compared to all of the other providers. This may be attributed to the amount of time it takes to treat children's mental health. 
```

```{r}
ggplot(df %>% 
         filter(los < 200) %>% 
         mutate(age_group = cut(age, breaks = c(0, 20, 40, 60, 80, 100), right = FALSE)), 
       aes(x = age_group, y = los, fill = factor(sex))) +
  geom_boxplot(outlier.color = "red", alpha = 0.7) +
  scale_fill_manual(values = c("blue", "pink"), labels = c("Male", "Female")) +
  labs(title = "Length of Stay (< 200 Days) by Gender and Age Group",x = "Age Group",y = "Length of Stay (LOS)",fill = "Gender",caption = "This Box plot shows LOS under 200 days grouped by gender and age ranges.") +
  theme_minimal()

# Here we can see that the length of stay for males and females for ages 0-20 are the highest, as the ages progress, males tend to have a longer length of stay. The high length of stay for the younger age group can most likely be attributed to women and infants and newborns, and also Bradley Hospital which had an average length of stay of almost 30 days. 
```

```{r}
avg_charges_by_gender <- df %>%
  group_by(sex) %>%
  summarize(avg_total = mean(total, na.rm = TRUE))

# Bar plot
ggplot(avg_charges_by_gender, aes(x = factor(sex), y = avg_total, fill = factor(sex))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("blue", "pink"), labels = c("Male", "Female")) +
  labs(title = "Average Total Charges by Gender",x = "Gender",y = "Average Total Charges",fill = "Gender",caption = "This Bar plot shows the average total charges for males and females.") +
  theme_minimal()

# Here we can see that males tend to have higher average total charges, compared to females. 
```

```{r}
library(gganimate)
df %>%
  group_by(moa, sex) %>%
  summarize(avg_los = mean(los, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = moa, y = avg_los, color = factor(sex))) +
  geom_line(size = 1.2) +  
  geom_point(size = 3) +  
  geom_text(aes(label = round(avg_los, 1)), vjust = -1, size = 4) + 
  scale_color_manual(values = c("blue", "pink"), labels = c("Male", "Female")) +
  labs(
    title = "Average Length of Stay In Different Months of the Year",
    subtitle = "Month of Admission: {frame_along}",
    x = "moa",
    y = "Average Length of Stay (LOS)",
    color = "Gender",
    caption = "This Animation shows trends in average LOS by gender in different months of the year."
  ) +
  theme_minimal() +
  transition_reveal(moa)

# This animation shows that during different months of the year males had a higher average length of stay than females. It is worth noting, that during the fall and winter months, male length of stays start to drop while female length of stays starts to rise. 
```
# Through visualizing the data analyzed in the first section, we were able to provide some meaningful plots that painted a good picture of what was going on in the data. The trends that were analyzed are total charges vs length of stay, total charges by age group, total charges by provider, the age distribution of patients, the gender distribution of patients, the distributions of charges by season, average length of stay by provider, length of stay by gender and age group, total charges by gender, and average length of stay in different months of the year by gender. Overall these plots allowed for us to find that Bradley Hospital had the highest average length of stay, patients are mostly newborns or between the ages of 55-80, and that men as well as younger age groups, have the highest total charges. One of the more prominent findings was that Rhode Island Hospital far outpaced the other hospital in total charges. 

# Predictive Models 
```{r}
library(tidyverse)
library(haven)
df2 = read_csv('midterm.csv')
n = nrow(df)
```

```{r}
df2 <- df2 %>%
  filter(yod == 18)
```

```{r}
library(dplyr)
df2 <- df2 %>%
  select("yod", "payfix","pay_ub92","age",  
                      "sex","raceethn","provider","moa", 
                      "yoa","mod","admtype", "asource" , 
                      "preopday" ,"los", "service" , "icu","ccu",    
                      "dispub92", "payer"  ,"drg","trandb", 
                      "randbg","randbs","orr", "anes","seq",   
                      "lab","dtest", "ther","blood","phar", 
                      "other","patcon","bwght","total","tot" ,  
                      "ecodub92","b_wt","pt_state","diag_adm","ancilar" ,
                      "campus","er_fee","er_chrg","er_mode","obs_chrg",
                      "obs_hour","psycchrg","nicu_day")
```

```{r}
missing_values <- df2 %>%
    summarise_all(~ sum(is.na(.)))

print(missing_values)
```
```{r}
df2 <- df2 %>%
  filter(!is.na(raceethn) & !is.na(admtype))
```


```{r}
# Creating Target variable
df2 <- df2 %>%
  mutate(target = ifelse(total < median(total, na.rm = TRUE), "low", "high"))

table(df2$target)
```

```{r}
# Input Variables
df2 <- df2 %>%
  select(age, sex, raceethn, provider, moa, mod, admtype, campus, los, target)
```

```{r}
# Use filter function to filter out rows where raceethn=='' or admtype==''. Make sure all the categorical variables are factor, numeric variables are numeric. Set Training : Testing Split = 10 : 90

library(tidyverse)
library(caret)
df2 <- df2 %>%
  filter(raceethn != '' & !is.na(raceethn) & admtype != '' & !is.na(admtype))

df2 <- df2 %>% 
  mutate(target = as.factor(target),
         raceethn = as.factor(raceethn),
         sex = as.factor(sex), 
         provider = as.factor(provider), 
         campus = as.factor(campus),
         admtype = as.factor(admtype), 
         moa = as.factor(moa), 
         mod = as.factor(mod))
splitIndex <- createDataPartition(df2$target, p = .10, 
                                  list = FALSE)
df2_train <- df2[ splitIndex,]
df2_test <- df2[-splitIndex,]


```

```{r}
# Train a decision tree using rpart. Plot the decision tree. Plot the variable importance ranked by the tree.

library(rpart)
tree1<-rpart(target ~ ., 
            data = df2_train,
            control=rpart.control(maxdepth=2))
# Plot the tree
library(rattle)
fancyRpartPlot(tree1)
```

```{r}
varImp(tree1)
```

```{r}
# Using caret for this question. Set Training Control to be: Use Cross-Validation of 5 folds across all models. Train & tune at least 2 different models (i.e. two different values for method= in the train function of caret). Plot the hyper-parameter tuning plots for each model.

tuneGrid = expand.grid(maxdepth = 2:10)
trControl = trainControl(method = "cv",
                         number = 5)
tree_approach2 <- train(target~., data=df2_train, 
                                method = "rpart2", 
                                trControl = trControl,
                                tuneGrid = tuneGrid)
```

```{r}
# Plot for first model
plot(tree_approach2)
```

```{r}
library(tidyverse)
library(caret)

trControl = trainControl(method = "cv",
                         number = 5)
tuneGrid = expand.grid(mtry = 3,
                       splitrule = c('gini', 'extratrees'),
                       min.node.size = c(10))
forest_ranger <- train(target~., data=df2_train, 
                    method = "ranger", 
                    trControl = trControl,
                    tuneGrid = tuneGrid)
```

```{r}
# Plot for second model
plot(forest_ranger)
```
```{r}
# Comparison of Models
trControl = trainControl(method = "cv",
                         number = 5)
tree <- train(target~., data=df2_train, 
                                method = "rpart2", 
                                trControl = trControl)
forest_ranger <- train(target~., data=df2_train, 
                    method = "ranger", 
                                trControl = trControl)
results <- resamples(list('Decision Tree' = tree,
                          'Random Forest' = forest_ranger))


bwplot(results)

```

```{r}
# Final Selection for the Model is the Random Forest model. 
library(caret)
pred <- predict(forest_ranger, df2_test)

cm <- confusionMatrix(data = pred, reference = df2_test$target, positive = "high")

cm$overall[1]
```
```{r}
df3 = read_csv('midterm.csv')
n3 = nrow(df3)
df3 <- df3 %>%
  filter(yod == 18)
```

```{r}
library(dplyr)
df3 <- df3 %>%
  select("yod", "payfix","pay_ub92","age",  
                      "sex","raceethn","provider","moa", 
                      "yoa","mod","admtype", "asource" , 
                      "preopday" ,"los", "service" , "icu","ccu",    
                      "dispub92", "payer"  ,"drg","trandb", 
                      "randbg","randbs","orr", "anes","seq",   
                      "lab","dtest", "ther","blood","phar", 
                      "other","patcon","bwght","total","tot" ,  
                      "ecodub92","b_wt","pt_state","diag_adm","ancilar" ,
                      "campus","er_fee","er_chrg","er_mode","obs_chrg",
                      "obs_hour","psycchrg","nicu_day")
```

```{r}
missing_values <- df3 %>%
    summarise_all(~ sum(is.na(.)))

print(missing_values)
```
```{r}
df3 <- df3 %>%
  filter(!is.na(raceethn) & !is.na(admtype))
```

```{r}
# Create Target Variable
df3 <- df3 %>%
  mutate(target = ifelse(age < 50 , "Young", "Old"))
table(df3$target)
    
```

```{r}
# Input Variables
df3 <- df3 %>%
  select(sex, raceethn, provider, moa, mod, admtype, campus,total,los, target)
```

```{r}
library(tidyverse)
library(caret)
df3 <- df3 %>%
  filter(raceethn != '' & !is.na(raceethn) & admtype != '' & !is.na(admtype))

df3 <- df3 %>% 
  mutate(target = as.factor(target),
         raceethn = as.factor(raceethn),
         sex = as.factor(sex), 
         provider = as.factor(provider), 
         campus = as.factor(campus),
         admtype = as.factor(admtype), 
         moa = as.factor(moa), 
         mod = as.factor(mod),
         total = as.numeric(total))
splitIndex <- createDataPartition(df3$target, p = .10, 
                                  list = FALSE)
df3_train <- df3[ splitIndex,]
df3_test <- df3[-splitIndex,]
```

```{r}
# Train a decision tree using rpart. Plot the decision tree. Plot the variable importance ranked by the tree.
library(rpart)
tree2<-rpart(target ~ ., 
            data = df3_train,
            control=rpart.control(maxdepth=2))
# Plot the tree
library(rattle)
fancyRpartPlot(tree2)

```

```{r}
varImp(tree2)
```
```{r}
# Using caret for this question. Set Training Control to be: Use Cross-Validation of 5 folds across all models. Train & tune at least 2 different models (i.e. two different values for method= in the train function of caret). Plot the hyper-parameter tuning plots for each model.

tuneGrid = expand.grid(maxdepth = 2:10)
trControl = trainControl(method = "cv",
                         number = 5)
tree_approach3 <- train(target~., data=df3_train, 
                                method = "rpart2", 
                                trControl = trControl,
                                tuneGrid = tuneGrid)

```

```{r}
# Plot for first model
plot(tree_approach3)
```

```{r}
# Second Model
library(tidyverse)
library(caret)

trControl = trainControl(method = "cv",
                         number = 5)
tuneGrid = expand.grid(mtry = 3,
                       splitrule = c('gini', 'extratrees'),
                       min.node.size = c(10))
forest_ranger2 <- train(target~., data=df3_train, 
                    method = "ranger", 
                    trControl = trControl,
                    tuneGrid = tuneGrid)
```

```{r}
plot(forest_ranger2)
```

```{r}
trControl = trainControl(method = "cv",
                         number = 5)
tree2 <- train(target~., data=df3_train, 
                                method = "rpart2", 
                                trControl = trControl)
forest_ranger2 <- train(target~., data=df3_train, 
                    method = "ranger", 
                                trControl = trControl)
results <- resamples(list('Decision Tree' = tree2,
                          'Random Forest' = forest_ranger2))


bwplot(results)
```

```{r}
# Final Selection is Random Forest Model

pred <- predict(forest_ranger2, df3_test)

cm <- confusionMatrix(data = pred, reference = df3_test$target, positive = "Old")

cm$overall[1]
```

# This section involved modeling the data, that was used to analyze trends and patterns in the first two sections. The first set of models, involved working with total charges, and creating a binary target variable. Total charges were seperated into a high and low category based on the median of total charges. The first tree created, using the total charges target, showed that a length of stay over 3.5 days indicated high total charges. If it was less than 3.5, it found that the admission type was usually electric or Truama. The next parts included cross validation, in which it was found that the Random Forest model was the most accurate one. The second tree was created using age as a target variable in which it was grouped into young for people less than 50 years old and old for people greater than 50 years old. It was found that a majority of the patients were older, and it was similar across all providers, and usually limited to Campus 0, 1, and 3, which are one campus hospitals, Fatima, or Rhode Island hospital. Like the first section, cross validation was used, and again found that Random Forest was more accurate in model prediction, than Decision Tree. 
